---
title: "Drug and Substance Abuse Treatment"
author: "Violet Allen-Glass, Isis Ramirez, Kaleigh Riggs, Jinchen Ye"
#output: pdf_document
output: ioslides_presentation
---

```{r, echo = FALSE, warning=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message = FALSE)

library(knitr) 
library(tidyverse) 
suppressMessages(library(tidyverse))
library(forcats) 
library(gridExtra) 
library(RSQLite) 
library(icd) 
library(jsonlite) 
library(gtable) 
library(grid) 
library(latex2exp) 
library(gridBase) 
library(nnet)
library(corrplot)
```

## Introduction

The data we are using comes from the Center for Behavioral Health Statistics and Quality Substance Abuse and Mental Health Services Administration.  It is their Treatment Episode Data Set- Admissions for 2016.  It includes 1,699,261 observations of 62 variables, each of these representing a person being admitted for substance abuse to a publicly funded facility.  Omitted from this data were the states Georgia and Oregon, because they did not report sufficient data in 2016.  

With these plots we are trying to see if there are any trends in services rendered or referral sources among the patient's employment status. We want to observe whether employment affects mental health and subsequent substance abuse.  Additionally, we want to find any trends in age by race to aid in identifying any racial disparity in access to mental health services.

```{r }
theme2 <- theme_bw() +
  theme(axis.text = element_text(size = 8, colour = "black"),
        axis.title = element_text(size = 10, colour = "black"), legend.title = element_text(size = 8, colour = "black"), legend.text = element_text(size = 8, colour = "black"), title = element_text(size = 12, colour = "black"), axis.ticks = element_line(colour = "black"),
        plot.caption = element_text(size = 8, colour = "black"), plot.subtitle = element_text(size = 10, colour = "black"))
tedsPalette <- c("seagreen3", "hotpink1", "darkorange1", "deepskyblue1", "mediumpurple2", "gold1", "red4", "grey60", "deeppink4", "orangered1")


# Main Data
dcon1 <- dbConnect(SQLite(), dbname = "TEDS.sqlite")

```

## Mental Health


## Accessibility to Treatment
```{r }
###################################################
###### Main Data ##############################
race <- dbSendQuery(conn = dcon1, "
SELECT RACE, DAYWAIT,
CASE 
            WHEN RACE ='1' THEN 'ALASKA NATIVE (ALEUT, ESKIMO, INDIAN)'
            WHEN RACE = '2'  THEN 'AMERICAN INDIAN (OTHER THAN ALASKA NATIVE)'
            WHEN RACE ='3' THEN 'ASIAN OR PACIFIC ISLANDER'
            WHEN RACE = '4'  THEN 'BLACK OR AFRICAN AMERICAN'
            WHEN RACE = '5'  THEN 'WHITE'
            WHEN RACE = '6'  THEN 'ASIAN'
            WHEN RACE = '7'  THEN 'OTHER SINGLE RACE'
            WHEN RACE = '8'  THEN 'TWO OR MORE RACES'
            WHEN RACE = '9'  THEN 'NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER '
            WHEN RACE = '-9' THEN 'MISSING/UNKNOWN/NOT COLLECTED/INVALID'
            ELSE NULL
       END AS RACE_DESCRIPTIONS
FROM tedsa_2016_puf;
")
df.main <- dbFetch(race, -1)

dbClearResult(race)

df.main <- df.main[df.main$DAYWAIT >= 0, ]

amerind <- filter(df.main, RACE == "1" | RACE == "2")
nam <- length(amerind$RACE)

amerind0 <- filter(amerind, DAYWAIT == "0")
am0 <- length(amerind0$RACE)/nam
amerind1 <- filter(amerind, DAYWAIT == "1")
am1 <- length(amerind1$RACE)/nam
amerind2 <- filter(amerind, DAYWAIT == "2")
am2 <- length(amerind2$RACE)/nam
amerind3 <- filter(amerind, DAYWAIT == "3")
am3 <- length(amerind3$RACE)/nam
amerind4 <- filter(amerind, DAYWAIT == "4")
am4 <- length(amerind4$RACE)/nam

asian <- filter(df.main, RACE == "3" | RACE == "6" | RACE == "9")
nas <- length(asian$RACE)
asian0 <- filter(asian, DAYWAIT == "0")
as0 <- length(asian0$RACE)/nas
asian1 <- filter(asian, DAYWAIT == "1")
as1 <- length(asian1$RACE)/nas
asian2 <- filter(asian, DAYWAIT == "2")
as2 <- length(asian2$RACE)/nas
asian3 <- filter(asian, DAYWAIT == "3")
as3 <- length(asian3$RACE)/nas
asian4 <- filter(asian, DAYWAIT == "4")
as4 <- length(asian4$RACE)/nas

black <- filter(df.main, RACE == "4")
nbla <- length(black$RACE)
black0 <- filter(black, DAYWAIT == "0")
b0 <- length(black0$RACE)/nbla
black1 <- filter(black, DAYWAIT == "1")
b1 <- length(black1$RACE)/nbla
black2 <- filter(black, DAYWAIT == "2")
b2 <- length(black2$RACE)/nbla
black3 <- filter(black, DAYWAIT == "3")
b3 <- length(black3$RACE)/nbla
black4 <- filter(black, DAYWAIT == "4")
b4 <- length(black4$RACE)/nbla

white <- filter(df.main, RACE == "5")
nwhi <- length(white$RACE)
white0 <- filter(white, DAYWAIT == "0")
w0 <- length(white0$RACE)/nwhi
white1 <- filter(white, DAYWAIT == "1")
w1 <- length(white1$RACE)/nwhi
white2 <- filter(white, DAYWAIT == "2")
w2 <- length(white2$RACE)/nwhi
white3 <- filter(white, DAYWAIT == "3")
w3 <- length(white3$RACE)/nwhi
white4 <- filter(white, DAYWAIT == "4")
w4 <- length(white4$RACE)/nwhi

other <- filter(df.main, RACE == "7" | RACE == "8")
noth <- length(other$RACE)
other0 <- filter(other, DAYWAIT == "0")
o0 <- length(other0$RACE)/noth
other1 <- filter(other, DAYWAIT == "1")
o1 <- length(other1$RACE)/noth
other2 <- filter(other, DAYWAIT == "2")
o2 <- length(other2$RACE)/noth
other3 <- filter(other, DAYWAIT == "3")
o3 <- length(other3$RACE)/noth
other4 <- filter(other, DAYWAIT == "4")
o4 <- length(other4$RACE)/noth
```

```{r}
race <- c("American Indian","Asian","Black","White","Other")
num1 <- c(am1,as1,b1,w1,o1)
num2 <- c(am2,as2,b2,w2,o2)
num3 <- c(am3,as3,b3,w3,o3)
num4 <- c(am4,as4,b4,w4,o4)
pc1 <- c('20.2%','27.5%','27.3%','33.5%','35.6%')
pc2 <- c('9.8%','10.5%','9.9%','12.6%','12.7%')
pc3 <- c('6.4%','6.3%','5.6%','7.2%','6.7%')
pc4 <- c('6.4%','6.3%','5.6%','7.2%','6.7%')
pc5 <- c('2.8%','2.8%','2.2%','3%','2.8%')
vp <- vector('list', 5)
vp[[1]] <- viewport(x = 0.25, y = 5/6, w = 0.5, h = 1/3)
vp[[2]] <- viewport(x = 0.25, y = 0.5, w = 0.5, h = 1/3)
vp[[3]] <- viewport(x = 0.75, y = 5/6, w = 0.5, h = 1/3)
vp[[4]] <- viewport(x = 0.75, y = 0.5, w = 0.5, h = 1/3)
vp[[5]] <- viewport(x = 0.5, y = 1/6, w = 0.5, h = 1/3)
grid.newpage()
for (i in 1:5) {
  pushViewport(vp[[i]])
  grid.rect(x = 0, y = 1, just = c("left", "top"), 
            height = 1/6, width = 6/7, gp = gpar(col = "black", lwd = 2, 
                                                 fill = tedsPalette[i]))
  grid.text(race[i], x = 0.05*6/7, y = 11/12, just = "left")
  grid.text("100%", x = 1.005*6/7, y = 0.85/6+5/6, just = "left")
  
  grid.rect(x = 0, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 2*6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 3*6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 4*6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 5*6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6*6/49, y = 4/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 0, y = 4/6, just = c("left", "bottom"), 
            height = (num1[i] + num2[i] + num3[i] + num4[i])/6, width = 6/7, 
            gp = gpar(col = tedsPalette[i], fill = tedsPalette[i]))
  grid.text("1", x = 0.005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("2", x = 6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("3", x = 2*6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("4", x = 3*6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("5", x = 4*6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("6", x = 5*6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text("7", x = 6*6/49 + .005*6/7, y = 0.85/6+4/6, just = "left")
  grid.text(pc1[i], x = 1.005*6/7, y = .2/6+4/6, just = "left")
  
  grid.rect(x = 0, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 2*6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 3*6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 4*6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 5*6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6*6/49, y = 3/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 0, y = 3/6, just = c("left", "bottom"), 
            height = (num2[i] + num3[i] + num4[i])/6, width = 6/7, 
            gp = gpar(col = tedsPalette[i], fill = tedsPalette[i]))
  grid.text("8", x = 0.005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("9", x = 6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("10", x = 2*6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("11", x = 3*6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("12", x = 4*6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("13", x = 5*6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text("14", x = 6*6/49 + .005*6/7, y = 0.85/6+3/6, just = "left")
  grid.text(pc2[i], x = 1.005*6/7, y = 0.1/6+3/6, just = "left")
  
  grid.rect(x = 0, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 2*6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 3*6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 4*6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 5*6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6*6/49, y = 2/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 0, y = 2/6, just = c("left", "bottom"), 
            height = (num3[i] + num4[i])/6, width = 6/7, 
            gp = gpar(col = tedsPalette[i], fill = tedsPalette[i]))
  grid.text("15", x = 0.005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("16", x = 6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("17", x = 2*6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("18", x = 3*6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("19", x = 4*6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("20", x = 5*6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text("21", x = 6*6/49 + .005*6/7, y = 0.85/6+2/6, just = "left")
  grid.text(pc3[i], x = 1.005*6/7, y = 0.065/6+2/6, just = "left")
  
  grid.rect(x = 0, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 2*6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 3*6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 4*6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 5*6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6*6/49, y = 1/6, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 0, y = 1/6, just = c("left", "bottom"), 
            height = (num3[i] + num4[i])/6, width = 6/7, 
            gp = gpar(col = tedsPalette[i], fill = tedsPalette[i]))
  grid.text("22", x = 0.005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("23", x = 6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("24", x = 2*6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("25", x = 3*6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("26", x = 4*6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("27", x = 5*6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text("28", x = 6*6/49 + .005*6/7, y = 0.85/6+1/6, just = "left")
  grid.text(pc4[i], x = 1.005*6/7, y = 0.065/6+1/6, just = "left")
  
  grid.rect(x = 0, y = 0, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 6/49, y = 0, just = c("left", "bottom"), height = 1/6,
            width = 6/49, gp = gpar(col = "black", lwd = 2))
  grid.rect(x = 0, y = 0, just = c("left", "bottom"), 
            height = (num3[i] + num4[i])/6,
            width = 2*6/49, gp = gpar(col = tedsPalette[i], 
                                      fill = tedsPalette[i]))
  grid.rect(x = 2*6/49, y = 0, just = c("left", "bottom"), height = num4[i]/6,
            width = 5*6/49, gp = gpar(col = tedsPalette[i], 
                                      fill = tedsPalette[i]))
  grid.text("29", x = 0.005*6/7, y = 0.85/6, just = "left")
  grid.text("30", x = 6/49 + .005*6/7, y = 0.85/6, just = "left")
  grid.text("30+", x = 2*6/49 + .005*6/7, y = 0.85/6, just = "left")
  grid.text(pc5[i], x = 1.005*6/7, y = .1/6, just = "left")
  popViewport()
}

```

## Hypothesis Testing
```{r}
testdata <- df.main %>% dplyr::select(RACE, DAYWAIT)

testdata$RACE[testdata$RACE=='2'] <- '1' 
testdata$RACE[testdata$RACE=='6'] <- '3'
testdata$RACE[testdata$RACE=='9'] <- '3'
testdata$RACE[testdata$RACE=='8'] <- '7'
table(testdata$RACE)

#Chi-square test examines whether rows and columns of a contingency
#table are statistically significantly associated.

#Null hypothesis (H0): the row and the column variables of 
#the contingency table are independent.
#Alternative hypothesis (H1): row and column variables are dependent

#Assumptions: 1.count data 2.mutually exclusive variables
#3.each subject contributes to only one cell
#4.independence 5.categorical or interval collapsed into ordinal data
#6.E(X) >= 5 for 80+%, E(X) > 1 for all

#For each cell of the table, we have to calculate the 
#expected value under null hypothesis

(chisq <- chisq.test(testdata$RACE, testdata$DAYWAIT))
chisq$observed
round(chisq$expected,2)


```
```{r}
#cells with the highest abs(Pearson, or standardized, residuals)
#contribute most to the total chi-square score
round(chisq$residuals,3)

```

## Correlation Plots  
```{r}
par(mfrow=c(1,2))
#visualizing pearson residuals. blue = positive association. 
#red = negative
#install.packages("corrplot")
library(corrplot)
corrplot(chisq$residuals, is.cor = FALSE)

#contribution (in %) of a given cell to the chi-square stat
contrib <- 100*chisq$residuals^2/chisq$statistic
#round(contrib, 3)
#visualizing the contribution
corrplot(contrib, is.cor = FALSE)
```

## Lifestyle 


## Drug Experience 
```{r}
comments.query <- dbSendQuery(conn = dcon1, "
SELECT *
FROM RedditCommentsQ1;
")
comments <- dbFetch(comments.query, -1)
dbClearResult(comments.query)

drug.labels <- c("Methamphetamine", "DMT", "Speed", "Coke")
meth.c <- sum(str_count(comments$Q1Comment, "meth|METH|Meth"))
dmt.c <- sum(str_count(comments$Q1Comment, "DMT"))
speed.c <- sum(str_count(comments$Q1Comment, "Speed|SPEED|speed"))
coke.c <- sum(str_count(comments$Q1Comment, "Coke|coke|COKE"))


drug.count <- rbind(meth.c, dmt.c, speed.c, coke.c)


meth.p <- sum(str_count(comments$Q1Comment, "meth|METH|Meth.*[incredible|joy|happy|positive]"))
dmt.p <-sum(str_count(comments$Q1Comment, "DMT.*[incredible|joy|happy|positive]"))
speed.p <-sum(str_count(comments$Q1Comment, "Speed|SPEED|speed.*[incredible|joy|happy|positive]"))
coke.p <-sum(str_count(comments$Q1Comment, "Coke|coke|COKE.*[incredible|joy|happy|positive]"))

drug.positive <- rbind(meth.p, dmt.p, speed.c, coke.p)


drug.positive.prop <- as.data.frame(drug.positive / drug.count)
col.names <- c("Drug", "Positive Words \n Proportion")
prop.table <- as.data.frame(cbind(drug.labels, format(round(drug.positive.prop, 2), nsmall = 2))) 

table <- tableGrob(prop.table, rows=NULL, cols = col.names)
#table <- grid.table(prop.table, rows=NULL)
#grid.arrange(table, gg, nrow=2)
```


```{r, fig.width=6, fig.height=10}
res <- dbSendQuery(conn = dcon1, "
SELECT RACE, SUB1
FROM tedsa_2016_puf;
")
data <- dbFetch(res, -1)
dbClearResult(res)
data$RACE[data$RACE=='1' | data$RACE=='2'] <- 'American Indian'
data$RACE[data$RACE=='3' | data$RACE=='6' | data$RACE=='9'] <- 'Asian'
data$RACE[data$RACE=='4'] <- 'Black'
data$RACE[data$RACE=='5'] <- 'White'
data$RACE[data$RACE=='7' | data$RACE=='8'] <- 'Other'
data <- data[data$RACE!='-9',]
data$SUB1[data$SUB1=='3'] <- 'COCAINE/CRACK'
data$SUB1[data$SUB1=='4'] <- 'MARIJUANA/HASHISH'
data$SUB1[data$SUB1=='5'] <- 'HEROIN'
data$SUB1[data$SUB1=='6'] <- 'NON-PRESCRIPTION METHADONE'
data$SUB1[data$SUB1=='7'] <- 'OTHER OPIATES AND SYNTHETICS'
data$SUB1[data$SUB1=='8'] <- 'PCP'
data$SUB1[data$SUB1=='9'] <- 'OTHER HALLUCINOGENS'
data$SUB1[data$SUB1=='10'] <- 'METHAMPHETAMINE'
data$SUB1[data$SUB1=='11'] <- 'OTHER AMPHETAMINES'
data$SUB1[data$SUB1=='12'] <- 'OTHER STIMULANTS'
data$SUB1[data$SUB1=='13'] <- 'BENZODIAZEPINES'
data$SUB1[data$SUB1=='14'] <- 'OTHER NON-BENZODIAZEPINE TRANQUILIZERS'
data$SUB1[data$SUB1=='15'] <- 'BARBITURATES'
data$SUB1[data$SUB1=='16'] <- 'OTHER NON-BARBITURATE SEDATIVES OR HYPNOTICS'
data$SUB1[data$SUB1=='17'] <- 'INHALANTS'
data$SUB1[data$SUB1=='18'] <- 'OVER-THE-COUNTER MEDICATIONS'
data <- data[data$SUB1!='1' & data$SUB1!='2' & data$SUB1!='19' & data$SUB1!='-9',]
names(data) <- c('Race', 'Substance')

data <- data %>% filter(Substance %in% c("COCAINE/CRACK", "MARIJUANA/HASHISH", "HEROIN", "METHAMPHETAMINE", "NON-PRESCRIPTION METHADONE"))
gg <- ggplot(data) +
  aes(x = Race, fill = Substance) +
  geom_bar(position = "fill") +
  labs(title = "Race against Substance Use",
       x = "Race", y = "Proportion of Substance Use") +
  theme2 +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

grid.arrange(table, gg, nrow=2)

```



```{r}
dbDisconnect(dcon1)
```

